---
title: "updated_data_exploration"
author: "Charles Hendrickson"
date: "2023-07-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(naniar)
library(visdat)
library(janitor)

#Turn off scientific notation as global setting
options(scipen=999)
```

Read in data
```{r}
lang_app_data <- read_csv(here("data/LangAppData.csv"))
```

Explore missing data
```{r}
vis_dat(lang_app_data)
```

Clean and prepare data
```{r}
#made column names lowercase
lang_app_data <- clean_names(lang_app_data)

#rename columns 
lang_app_data <- lang_app_data %>%
         rename("household_language" = "lanp_label",
                "state" = "st_label",
                "speak_english" = "eng_label",
                "speakers" = "n")

#create new column for state abbreviations
lang_app_data <- lang_app_data %>% separate(col = state, into = c("state", "state_abb"), sep = "/")

#check data type of columns
summary(lang_app_data)

#Remove 'state_abb' column because it is unnecessary
lang_app_data <- lang_app_data %>% select(!state_abb)
```

Get stats on languages and make new columns
```{r}
#Create new column of the total number of speakers of each language
lang_app_data <- lang_app_data %>% 
                 group_by(household_language) %>% 
                 mutate(total_speakers_of_language_in_us = sum(speakers))

#Find the percent of speakers for each state
lang_app_data <- lang_app_data %>% mutate(percentage_speakers_in_state = speakers/total_speakers_of_language_in_us*100)

#Find the population of individuals with household language that is not English 
lang_app_data <- lang_app_data %>% group_by(state) %>% mutate(state_pop_of_non_english_households = sum(speakers))

#Find Percent of State Pop. Speaking Language
lang_app_data <- lang_app_data %>% mutate(percent_state_pop_speaking_language = speakers/state_pop_of_non_english_households*100)
```


Sanity Check: Explore data on a specific Language in a State
```{r}
#Explore Spanish in California
#test <- lang_app_data %>% filter(household_language == "Spanish" & state == "California")
```


Round to two decimal places
```{r}
#Round percentages to two decimal places
lang_app_data <- lang_app_data %>% mutate(across(where(is.numeric), ~ round(., digits = 4)))
```

This function rounds numbers less than 1 to two decimal places after the first one-digit natural number, and rounds numbers greater than or equal to 1 to two decimal places. 

(*This is so that values smaller than 0.00 are not rounded to 0, which would make the user think no people speak the language.)
```{r}
#Function: custom_round

#Description: Rounds numbers less than 1 to two decimal places after the first one-digit natural number,
              #and rounds numbers greater than or equal to 1 to two decimal places.

# Arguments:
#   x: A numeric vector containing the values to be rounded.
#
# Returns:
#   A numeric vector with the rounded values.
#
# Example:
#   custom_round(c(0.12345, 0.98765, 1.234567, 9.876543))
#   Output: 0.13 0.99 1.23 9.88

# custom_round <- function(x) {
#   rounded_values <- ifelse(x < 1,
#                            round(x, digits = 2 - floor(log10(abs(x)))),
#                            round(x, digits = 2))
#   return(rounded_values)
# }
```

Test the function above
```{r}
#test <- data.frame(custom_round(lang_app_data$percentage_speakers_in_state))
```


```{r}
#test <- lang_app_data %>% filter(household_language == "Spanish",
                                 #state == "California",
                                 #speak_english == "Well")
```


```{r}
#Add commas to thousandth place in speakers column
lang_app_data <- lang_app_data %>%
  mutate(speakers = format(speakers, big.mark = ","))
```

```{r}
#Add commas to thousandth place in total_speakers_of_language_in_us column
lang_app_data <- lang_app_data %>%
  mutate(total_speakers_of_language_in_us = format(total_speakers_of_language_in_us, big.mark = ","))
```

```{r}
#Add commas to thousandth place in state_pop_of_non_english_households column
lang_app_data <- lang_app_data %>%
  mutate(state_pop_of_non_english_households = format(state_pop_of_non_english_households, big.mark = ","))
```

Add state abbreviation
```{r}
# Create a data frame with state names and abbreviations
state_abbreviations <- data.frame(
  state = c("Alaska", "Alabama", "Arkansas", "Arizona", "California", "Colorado",
             "Connecticut", "District of Columbia", "Delaware", "Florida", "Georgia",
             "Hawaii", "Iowa", "Idaho", "Illinois", "Indiana", "Kansas", "Kentucky",
             "Louisiana", "Massachusetts", "Maryland", "Maine", "Michigan", "Minnesota",
             "Missouri", "Mississippi", "Montana", "North Carolina", "North Dakota",
             "Nebraska", "New Hampshire", "New Jersey", "New Mexico", "Nevada",
             "New York", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island",
             "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Virginia",
             "Vermont", "Washington", "Wisconsin", "West Virginia", "Wyoming"),
  state_abbreviation = c("AK", "AL", "AR", "AZ", "CA", "CO", "CT", "DC", "DE", "FL",
                         "GA", "HI", "IA", "ID", "IL", "IN", "KS", "KY", "LA", "MA",
                         "MD", "ME", "MI", "MN", "MO", "MS", "MT", "NC", "ND", "NE",
                         "NH", "NJ", "NM", "NV", "NY", "OH", "OK", "OR", "PA", "RI",
                         "SC", "SD", "TN", "TX", "UT", "VA", "VT", "WA", "WI", "WV", "WY")
)

# Assuming "census_data" is your existing data frame
lang_app_data <- lang_app_data %>%
  left_join(state_abbreviations, by = "state")

```


```{r}
#Create CSV of dataframe 
#write.csv(lang_app_data,
          #"lang_app_data.csv", row.names=FALSE)
```













